---
title: "RNAsep2 De analysis report"
author: "Thomas Sol Dourdin"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Differential gene expression analysis using DESeq2**

| The differential gene expression analysis was performed using the DESeq2 R package.

| The three treated conditions: **Hg**, **CO2** and **Hg+CO2** were independently analysed against the control condition: **CT**, using the *Wald test* implemented in DESeq2 (pairwise comparison).
The threshold used were : **pvalue < 0.05** & **|log2FoldChange| > 0.5**.

| Regarding the **Hg** & **CO2** conditions, only few genes were differentially expressed : **5** and **15**, respectively. However, **315 DEGs** were found in the **Hg+CO2** condition, suggesting a synergistic interaction between both stressors when they occur concomitantly.  

| As shown in the Venn diagram below, only one DEG was found to be common to all conditions but a substantial proportion of DEGs from **Hg** and **CO2** conditions were also affected by the **Hg+CO2** treatment (60% and 27%, respectively).


```{r echo=FALSE, warning=FALSE, message=FALSE}
grid.newpage()
venn_object <- tar_read(VennDiagram)
grid.draw(venn_object)
```


| Among the 315 DEGs from the **Hg+CO2** condition, **61 were up-regulated** and **254 were down-regulated** (cf. heatmap).
|
|
|

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=6}
tar_read(Heat1)
```


# **Function enrichment analysis using GOMWU**

| Overall, 158 biological processes were found to be significantly enriched in the **Hg+CO2** condition. The more significants are represented below:
|
|
|

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
tar_read(HgCO2_BP_Plot)
```

| Noteworthy, among the main repressed GOterms are biological processes related to the defense of the organism (i.e. $*immune response*, *defense response*, *response to wounding*, *inflammatory response*, *protein activation cascade* etc.). A link to the alteration of anti-predatory behaviour may be done. Moreover, genes implicated in the development of musclar tissues (i.e. *muscle development* and *muscle cell differentiation*), along with the development of neuronal network (i.e. *defasciculation of motor neuron axon*, *axonal defasciculation*) are also repressed, corroborating behavioral evidences for the neurotixicity of the two stressors.

| On the other side, there is a clear induction of RNA processing-related genes (*ribosome biogenesis*, *RNA splicing*, *mRNA processing*) as well as putative epigenetic regulation (*chromatin organization*, *obsolete histone H3-K4 methylation*, *obsolete covalent chromatin modification*) which may be characteristic of a response to environmental stress. Based on this, investigating epigenetic modifications would be  quite relevant. (**See also the growing literature on the topic**).
